---
title: "fMRI utilities"
author: "Eric Bridgeford"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beta Graph Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  header-includes:
   - \usepackage{amsfonts}
   - \usepackage{amsmath}
   - \usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

In this page, we will cover much of the basic functionality of the `fmriutils` package, including loading, processing, visualizing, and manipulating data.

# Loading Data

## BIDS Spec

When working with data, it is often advantageous to apply a standard specification, including directory organization, naming convention, required data, etc. For the purposes of simplicity, we will use the [BIDs specification](http://bids.neuroimaging.io/), a spec designed for magnetic-resonance derivatives of multiple modalities. The general format of files is: `sub-[####]_task-[abcd...]_ses-[####]_detailed_info.ftype`

In this section, we will cover the basics of loading data. For this tutorial, we assume that we have the following output structure produced by our pipeline:

```{r, eval=FALSE}
+ outputs/
    + BNU1/
        + graphs/
            + desikan-2mm/  # atlas name
                + sub-0025864_ses-1_desikan-2mm.graphml
                + ...
            + aal-2mm/  # atlas name
                + sub-0025864_ses-1_aal-2mm.graphml
                + ...
        + timeseries/
            + desikan-2mm/
                + sub-0025864_ses-1_desikan-2mm.rds
                + ...
            + aal-2mm/
                + sub-0025864_ses-1_aal-2mm.rds
                + ...
    + DC1/
        + graphs/
            + desikan-2mm/  # atlas name
                + sub-0025516_ses-1_desikan-2mm.graphml
                + ...
            + aal-2mm/  # atlas name
                + sub-0025516_ses-1_aal-2mm.graphml
                + ...
        + timeseries/
            + desikan-2mm/
                + sub-0025516_ses-1_desikan-2mm.rds
                + ...
            + aal-2mm/
                + sub-0025516_ses-1_aal-2mm.rds
                + ...
```

That is, we have our directories structured such that all of our timeseries/graphs for a single dataset, single atlas are in a single folder. This is not a "hard requirement" for using the fmriutils package, but it will make our life much easier down the line, and will allow us to easily aggregate dataset and atlas-level identification information associated with our data. Assuming that we are in the same directory as the `outputs/` folder, we can load our data as follows for a single dataset, single atlas:

```{r, eval=FALSE}
require(fmriutils)
# the path to our graphs or timeseries
sig_inpath <- './outputs/BNU1/timeseries/desikan-2mm/'
gra_inpath <- './outputs/BNU1/graphs/desikan-2mm/'

# get vectors where each element is a file name
sig_names <- list.files(sig_inpath, pattern="\\.rds", full.names=TRUE)
gra_names <- list.files(gra_inpath, pattern="\\.graphml", full.names=TRUE)

# open up the rds timeseries files
signalobj <- open_timeseries(sig_names, dataset_id = 'BNU1', atlas_id = 'desikan-2mm')

graphobj <- open_timeseries(gra_names, dataset_id = 'BNU1', atlas_id = 'desikan-2mm')
```

If we have multiple datasets, and multiple atlases, we can easily aggregate using this structuring of inputs. For example, in the example above, we may want to aggregate over the BNU1 and DC1 datasets, and aggregate over the desikan-2mm and aal-2mm atlases:

```{r, eval=FALSE}
datasets <- c('BNU1', 'DC1')
atlases <- c('desikan-2mm', 'aal-2mm')

# initialize data structures to save the derivatives and names
timeseries <- list()
subject_ids <- c()
session_ids <- c()
dataset_ids <- c()
task_ids <- c()

# iterate over datasets
for (dataset in datasets) {
  # iterate over atlases
  for (atlas in atlases) {
    # aggregate the timeseries names
    sig_names <- list.files(sig_inpath, pattern="\\.rds",
                            full.names=TRUE)

    # open the timeseries we have data for
    signalobj <- open_timeseries(sig_names, dataset_id = dataset,
                                 atlas_id = atlas, sub_pos = 2,
                                 ses_pos=4, exclude = FALSE)
    # append outputs to our initial data structures
    timeseries <- append(timeseries, signalobj$ts)
    dataset_ids <- c(dataset_ids, signalobj$dataset)
    atlas_ids <- c(atlas_ids, signalobj$atlas)
    subject_ids <- c(subject_ids, signalobj$subjects)
    session_ids <- c(session_ids, signalobj$sessions)
    task_ids <- c(task_ids, signalobj$tasks)
  }
}
```

and our resulting timeseries list, and vectors for subject/session/task/dataset level parameters, will be organized appropriately. NOTE: when attempting to aggregate outputs as was just described, make sure your filenames are descriptive enough such that no two files that you add have the same filename character for character. For example, in the examples above, we have multiple versions of subject 0025864 session 1 timeseries data (1 for desikan-2mm and 1 for aal-2mm), but since each timeseries is uniquely identified with the atlas name, we will not have any overlap. 

# Processing Data

Moreover, as a user it is often valuable to be able to manipulate your data into other desirable formats. 

# Visualizing Data

# Manipulating Data
