---
title: "fMRI utilities"
author: "Eric Bridgefordr"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beta Graph Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  header-includes:
   - \usepackage{amsfonts}
   - \usepackage{amsmath}
   - \usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
---

In this page, we will cover much of the basic functionality of the `fmriutils` package, including loading, processing, visualizing, and manipulating data.

# Loading Data

In this section, we will cover the basics of loading data. For this tutorial, we assume that we have the following output structure produced by our pipeline:

```{eval=FALSE}
+ outputs/
    + BNU1/
        + graphs/
            + desikan-2mm/  # atlas name
                + sub-0025864_ses-1_desikan-2mm.graphml
                + ...
            + aal-2mm/  # atlas name
                + sub-0025864_ses-1_aal-2mm.graphml
                + ...
        + timeseries/
            + desikan-2mm/
                + sub-0025864_ses-1_desikan-2mm.rds
                + ...
            + aal-2mm/
                + sub-0025864_ses-1_aal-2mm.rds
                + ...
    + DC1/
        + graphs/
            + desikan-2mm/  # atlas name
                + sub-0025516_ses-1_desikan-2mm.graphml
                + ...
            + aal-2mm/  # atlas name
                + sub-0025516_ses-1_aal-2mm.graphml
                + ...
        + timeseries/
            + desikan-2mm/
                + sub-0025516_ses-1_desikan-2mm.rds
                + ...
            + aal-2mm/
                + sub-0025516_ses-1_aal-2mm.rds
                + ...
```

That is, we have our directories structured such that all of our timeseries/graphs for a single dataset, single atlas are in a single folder. This is not a "hard requirement" for using the fmriutils package, but it will make our life much easier down the line, and will allow us to easily aggregate dataset and atlas-level identification information associated with our data. Assuming that we are in the same directory as the `outputs/` folder, we can load our data as follows for a single dataset, single atlas:

```{r}
# the path to our graphs or timeseries
sig_inpath <- './outputs/BNU1/timeseries/desikan-2mm/'
gra_inpath <- './outputs/BNU1/graphs/desikan-2mm/'

# get vectors where each element is a file name
sig_names <- list.files(sig_inpath, pattern="\\.rds", full.names=TRUE)
gra_names <- list.files(gra_inpath, pattern="\\.graphml", full.names=TRUE)

# open up the rds timeseries files
# note that our file-ids are delimimted by any special characters
# ., -, or _, so or sub id is at 2 and our session id at 4
signalobj <- open_timeseries(sig_names, dataset_id = 'BNU1', atlas_id = 'desikan-2mm', sub_pos = 2, ses_pos=4, exclude = FALSE)

graphobj <- open_timeseries(gra_names, dataset_id = 'BNU1', atlas_id = 'desikan-2mm', sub_pos = 2, ses_pos=4, exclude = FALSE)
```

If we have multiple datasets, and multiple atlases, we can easily aggregate using this structuring of inputs:

```{r}
datasets <- c('BNU1', 'DC1')
atlases <- c('desikan-2mm', 'aal-2mm')

# initialize data structures to save the derivatives and names
timeseries <- list()
subject_ids <- c()
session_ids <- c()
dataset_ids <- c()

# iterate over datasets
for (dataset in datasets) {
  # iterate over atlases
  for (atlas in atlases) {
    # aggregate the timeseries names
    sig_names <- list.files(sig_inpath, pattern="\\.rds",
                            full.names=TRUE)

    # open the timeseries we have data for
    signalobj <- open_timeseries(sig_names, dataset_id = dataset,
                                 atlas_id = atlas, sub_pos = 2,
                                 ses_pos=4, exclude = FALSE)
    # append outputs to our initial data structures
    timeseries <- append(timeseries, signalobj$ts)
    dataset_ids <- c(dataset_ids, signalobj$dataset)
    atlas_ids <- c(atlas_ids, signalobj$atlas)
    subject_ids <- c(subject_ids, signalobj$subjects)
    session_ids <- c(session_ids, signalobj$sessions)
  }
}
```

# Processing Data

# Visualizing Data

# Manipulating Data
